Kinetic Modeling of Isobutane Alkylation with Mixed C4 Olefins and Sulfuric Acid as a Catalyst Using the Asynchronous Global Optimization Algorithm

### 2022-04-30 2022/PAVT/PAVT2022.tex
final
after proofreading

\begin{document}
\mainmatter              % start of a contribution
% Решение обратных задач химической кинетки с помощью асинхронного алгоритма глобального поиска
% Решение обратных задач химической кинетки с помощью смешанного локально-глобального поискового алгоритма 
%\title{Solving the Inverse Problems of Chemical Kinetics Using the Asynchronous Global Optimization Algorithm}
\title{Kinetic Modeling of Isobutane Alkylation with Mixed C4 Olefins and Sulfuric Acid as Catalyst Using the Asynchronous Global Optimization Algorithm}
\titlerunning{Kinetic Modeling of Isobutane Alkylation}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{
Irek Gubaydullin$^{1,2}$\and
Leniza Enikeeva$^{2,3}$\orcidID{0000-0003-4219-4870}
\and
Konstantin Barkalov$^4$ \letter \orcidID{0000-0001-5273-2471}
\and
Ilya Lebedev$^4$\orcidID{0000-0002-8736-0652} 
\and
Dmitry Silenko$^4$ 
}

@@

\institute{$^1$Institute of Petrochemistry and Catalysis – Subdivision of the Ufa Federal Research Centre of RAS, Ufa, Russia\\$^2$Ufa State Petroleum Technological University, Ufa, Russia \\$^3$Novosibirsk State University, Novosibirsk, Russia\\$^4$Lobachevsky State University of Nizhny Novgorod, Nizhny Novgorod, Russia\\
\email{leniza.enikeeva@yandex.ru},
\email{konstantin.barkalov@itmm.unn.ru},
\email{ilya.lebedev@itmm.unn.ru}
}

@@

The paper considers the application of parallel computing technology to the simulation of catalytic chemical reaction, which is widely used in the modern automobile industry to produce gasoline with high octane number. As a chemical reaction, the process of alkylation of isobutane with mixed C4 olefins catalyzed by sulfuric acid is assumed. To simulate a chemical process, it is necessary to develop a kinetic model of the process, that is, to determine the kinetic parameters. To do this, the inverse problem of chemical kinetics is solved, which predicts the values of kinetic parameters based on laboratory data. From a mathematical point of view, the inverse problem of chemical kinetics is a global optimization problem. A parallel asynchronous information-statistical global search algorithm was used to solve it. The use of the asynchronous algorithm has significantly reduced the search time to find the optimum. The found optimal parameters of the model made it possible to adequately simulate the process of alkylation of isobutane with mixed C4 olefins catalyzed by sulfuric acid.

%обновить ключевые слова
\keywords{Global optimization $\cdot$ Multiextremal functions $\cdot$ Parallel computing $\cdot$ Chemical kinetics $\cdot$ Inverse problems }
\end{abstract}

@@

%Часть УГНТУ
Currently, there is a tendency to improve the environmental characteristics of automobile fuel while maintaining a high octane number. Sulfuric acid alkylation of isobutane with olefins makes it possible to obtain a high-octane component of gasoline with a minimum content of aromatic hydrocarbons. The alkylate, which is produced by the alkylation of isobutane with C3 -- C5 olefins in the presence of strong acid, has the advantages of high octane number, low vapor pressure, and zero content of olefins and aromatics that allow it to be a desirable blending component for high-quality gasoline. Alkylates will continue to act as a desirable blending component for high-quality gasoline as the quality of gasoline continues to increase \cite{cao2019}. Therefore, it is a significant process of the modern refinery. To optimize the chemical process in industry, it is necessary to first develop its model, which in this case means building a mathematical model of the chemical process and then its kinetic model, that is, numerically calculate the kinetic constants of the reaction.

%Стыковочный абзац
As a rule, it is impossible to find out the kinetic constants of the reactions analytically. Therefore, there is a need in the development and application of numerical methods for finding the kinetic constants (see, e.g., \cite{Zaynullin2020,Enikeeva2020,Koledina2019,Enikeev2020,uskovfibrous,enikeeva2021,en13133393}). In this case, the quality criteria of the solution found (objective function) haven't an explicitt analytical description but allows an algorithmic representation and requires considerable computaton resources. Moreover, in the inverse problems of chemical kinetics the objective function can be multiextremal essentially, i.e. can have many local extrema along with the global one.

%Часть ННГУ
The numerical methods for solving such multiextremal problems (global optimization methods) differ significantly from the local search ones (see, e.g., \cite{Sergeyev2017,PaulaviciusZilinskas2014}). As a rule, the local optimization methods cannot escape a local extremum attractione region and don't find the global optimum. At the same time, the use of the model parameters corresponding to the local solution found may appear to be insufficient since the global solution may provide a considerable advantage as compared to the local ones.

The diversity of the global optimization problems arising entails various approaches to solving these ones. The methods of solving the global optimization problems can be divided into two classes: the metaheuristic methods and the deterministic ones. The metaheuristic algorithms, as a rule, are based on the simulation of the processes going in the nature. Some examples of the metaheuristic algorithms are simulated annealing, evolution and genetic algorithms, etc. (see e.g. \cite{Battiti2009,Eiben2015}). Because of relative simplicity, the metaheuristic algorithms are more popular among the researchers than the deterministic methods. 
However, a problem solution found by a metaheuristic algorithm is, generally speaking, a local one and may be far from the global solution \cite{Kvasov2018}.

The possibility to construct deterministic global search methods different from the grid search and from the metaheuristic methods is related to the availability of and taking into account some {\it a priori} assumption on the properties of the problem functions. Such assumptions play a key role in the development of efficient global optimization algorithms and serve as main mathematical tools for estimating the global solutions.

An assumption on limited relative variations of the objective function values is one of natural assumptions of a problem. Such an assumption is related to the ratio of the function increment to respective increment of its argument, which is usually limited by some threshold defined by a limited energy of variations in the simulated system. In this case the functions are known as Lipschitz ones and the problem itself is called the Lipschitz global optimization problem.

This paper presents the results of application of parallel Lipschitz optimization methods for solving the inverse problems of chemical kinetics. The main part of the paper has the following structure. The description of the mathematical model of the investigated chemical reaction is presented in Section \ref{Sec_math_mod}. The formal statement of the Lipschitz global optimization problems and the asynchronous parallel algorithm for solving these ones are described in Section \ref{Sec_GSA}. The results of the numerical solving the inverse problem of chemical kinetics are discussed in Section \ref{Sec_Exp}.

@@

The initial conditions are $t = 0, c_1 = c_1^0; c_2=c_2^0; c_3 = 0; c_4 = 0; c_5= 0; c_6 = 0; c_7 = 0; c_8 = 0; c_9 = 0; c_{10} = 0; c_{11}=c_{11}^0; c_{12} = c_{12}^0$.
The corresponding species in eqs (\ref{eq:one})--(\ref{eq:twelve}) are 1, iC4H8; 2, iC4, 3, iC4+; 4, TMPs+; 5, DMHs+; 6, TMPs; 7, DMHs; 8, HEs; 9, iCx+; 10, iCy=; 11, 2-C4H8; 12, 1-C4H8.

As experimental data, information is taken from the article \cite{cao2019}, which represent a change in the concentrations of reaction components over time at different temperatures; as an example of experimental data at a temperature of 276.2 K, the data in Table \ref{table1} are presented.

\begin{table}
\caption{Experimental Data}
\label{table1}
\begin{center}
\begin{tabular}{ccccccc}
\hline
time, min & 1 & 2 & 5 & 10 & 15 & 20 \\
\hline\rule{0pt}{12pt}
DMH & 0.12 & 0.11 & 0.1	& 0.1 &	0.095 &	0.09  \\
TMP & 0.54 & 0.65 & 0.69 & 0.69 & 0.7 & 0.705 \\[2pt]
\hline
\end{tabular}
\end{center}
\end{table}

@@

However, it is necessary to take into account the fact that the reaction rate constants $k_1, k_2, ..., k_{15}$ included in the equations (\ref{eq:one})--(\ref{eq:twelve}) are parameters depending on the reaction temperature, and this dependence is the Arrhenius equation and has the following form:

\begin{equation}
  k_i (T) = k_i^0 \exp \left(- \dfrac{E_i}{RT} \right),
  \label{eq:arren}
\end{equation}
where $k_i(T)$ is a constant of $i$-th stage of reation rate, $k_i^0$ is a pre-exponential factor of $i$-th reation stage, $E_i$ is an activation energy, J/mol, $R$ is the universal gas constant, J $\cdot$ (K$\cdot$mol), $T$ is the temperature, K. Thus, in order to fully develop the kinetic model of the reaction, it is necessary to calculate the activation energies $E_i$ and pre-exponential factors $k_i^0$ of all stages of the chemical reaction.  There are two formulations of the problems of searching for kinetic parameter data $E$ and $k^0$. The first one is to solve the inverse problem of selecting the kinetic parameters included in the equation (\ref{eq:arren}), which during the solution allow us to calculate all the reaction rate constants $k_i$. Then, with the found rate constants of the reaction stages, the system of differential equations (\ref{eq:one})--(\ref{eq:twelve}) is solved, then the calculated concentrations are compared with the corresponding experimental data. Mathematically, this problem has the following formulation: it is necessary to minimize the following objective function

\begin{equation}
  F_1 = \sum_{i=1}^I \sum_{j=1}^J \sum_{k=1}^K \left| c_{ijk}^{exp} - c_{ijk}^{calc} \right| \longrightarrow \min,
  \label{eq:objective_func1}
\end{equation}
where $c_{ijk}^{exp}$ and $c_{ijk}^{calc}$ are the experimental and calculated values of $k$-th observed component at $i$-th experiment, respectively, $I$ is the the number of observed temperatures for the reaction, ($I = 4$ in this case), $K$ is the number of experiments conducted at one temperature ($K = 3$), $M$ is the number of observed components of the reaction ($M = 2$).

The second formulation of the inverse problem of chemical kinetics implies the search for the rate constants of the stages $k_i$ included in the system (\ref{eq:one})--(\ref{eq:twelve}), separately for each temperature, then on the basis of the Arrhenius equation~(\ref{eq:arren}), the activation energies $E_i$ and the pre-exponential multipliers $k_i^0$ are calculated using the least squares method. Mathematically, this statement of the problem coincides with the equation (\ref{eq:objective_func1}), except for the first summation by temperatures:

@@

Thus, both formulations of the problem (\ref{eq:objective_func1}) and (\ref{eq:objective_func2}) are optimization problems and the next section will describe the method that was used to solve these minimization problems.

@@

As it has been already mentioned above, from the formal point of view we consider the inverse problem of chemical kinetics as a global optimization problem. 
In the specific problem under consideration, the values of the objective function are calculated by solving stiff ODE system (\ref{eq:one})--(\ref{eq:twelve}). Since the right parts of the system are continuous functions with bounded derivatives, then theoretically its solution will also be continuous and bounded. Therefore, the discrepancy (\ref{eq:objective_func2}) will satisfy the Lipschitz condition with a priori unknown constant.

In the general form, a problem of the class specified above can be formulated mathematically as 
follows:
\begin{gather}
 \varphi^* = \varphi(y^\ast)=\min{\left\{\varphi(y):y\in D\right\}}, \label{problemN}\\
 D=\left\{y\in R^N: a_i\leq y_i \leq b_i, \;  1\leq i \leq N\right\} \label{D},
\end{gather}
where $a,b$ are given vectors, $a,b\in R^N$, and the objective function $\varphi(y)$ satisfies the Lipschitz condition
\begin{equation}\label{Lip}
\left|\varphi(y_1)-\varphi(y_2)\right|\leq L\left\|y_1-y_2\right\|,\; y_1,y_2 \in D.
\end{equation}

The function $\varphi(y)$ is assumed to be multiextremal and defined in the form of ``black box'' (i.e. in the form of some computing procedure, the input of which the  vector of parameters is supplied into, and the corresponding function value is taken from the output). Moreover, each \textit{trial} (i.e. the computing of the function value at a point of the search domain) is assumed to be time-consuming operation. 
As it has been noted in Introduction, such problem statement corresponds to the inverse problem of chemical kinetics completely.

The Lipschitz condition (\ref{Lip}) can be utilized to estimate the global minimum of a function within an interval, and knowing the Lipschitz constant allows constructing the global search algorithms and proving the convergence conditions for these ones (see, for example, \cite{Strongin2000}).

The growth of the computation costs with increasing problem dimensionality is one of the main difficulties in solving the multidimensional global optimization problems. The decreasing of the number of trials at preserving the solution accuracy is possible by a complete utilization of some {\it a priori} assumptions on the objective function that leads to adaptive sequential optimization algorithms.

For example, non-uniform space covering method \cite{Evtushenko2013} and simplicial partitions method \cite{Zilinskas2010} are such methods. These approaches were applied successfully for the development of the parallel optimization methods as well \cite{Evtushenko2009,Paulavicius2011}. 
Another adaptive approach to solving a multidimensional problem (\ref{problemN}) is its reduction to a single one-dimensional problem or to several ones followed by application of the one-dimensional algorithms. 
Such a reduction can be made, for example, using nested optimization scheme \cite{Grishagin2018} or Peano-Hilbert curves \cite{Barkalov2018}. 
The latter approach was used in the present work.

Using the continuous unambiguous mapping (Peano-Hilbert curve) $y(x)$ of the interval $[0,1]$ of the real axis on the hypercube $D$ from (\ref{D}), one can reduce a multidimensional problem (\ref{problemN}) to a one-dimensional problem
\[
\varphi(y^\ast)=\varphi(y(x^\ast))=\min{\left\{\varphi(y(x)): x\in[0,1]\right\}},
\]
where the function $\varphi(y(x))$ will satisfy a uniform H{\"o}lder condition
\[
\left|\varphi(y(x_1))-\varphi(y(x_2))\right|\leq H\left|x_1-x_2\right|^{1/N}
\]
with the H{\"o}lder constant $H$ linked to the Lipschitz constant $L$ by the relation $ H=2 L \sqrt{N+3}$. 
The issues of the numerical construction of various approximations of the Peano-Hilbert curve were considered in \cite{Strongin2000,Sergeyev2013}.

So far, a search trial at some point $x'\in[0,1]$ will include first the construction of the image $y'=y(x')$ and then the computing the value of the function $z'=\varphi(y')$.

@@

In the approach proposed, the parallelization scheme corresponds to the ``master/worker'' principle.
In the master process the global search algorithm is executed, which accumulates the search information, evaluates the Lipschitz constant for the objective function on its base, determines the new trial points and distributes these ones among the worker processes.

The worker processes receive the trial points from the master process, perform the new trials at these points, and send the trial results to the master process.

Let us assume the master process to compute one point of the next trial at every iteration and to send it to a worker process for executing the trial. 
At the same time, the execution of the trial by the worker process is much more computational-costly operation than the choice of a new trial point by the master that excludes idle worker processes. 
In this case (unlike the synchronous parallel algorithms), the total number of trials executed by each worker process will depend on the computation costs of executing a particular trial and cannot be estimated in advance.

In the description of the parallel algorithm, let us assume $p+1$ computing processes to be at our disposal: one master process and $p$ worker ones.

At the beginning of the search, the master process (let us assume it to be Process No 0) initiates the parallel execution of $p$ trials at $p$ different points of the search domain. 
Two of these points are the boundary ones while the rest are the internal ones, i.e. at the points $\{y(x^1), y(x^2), ...,y(x^p)\}$ where 
$x^1 = 0$, $x^p = 1$, $x^i\in(0,1), i=2,..., p-1$.

Now let us assume $k\geq 0$ trials (in particular, $k$ may be equal to 0) to be completed, and the worker processes are performing the trials at the points $\{y(x^{k+1}), y(x^{k+2}), ...,y(x^{k+p})\}$.

Each worker process having completed its trial at some point (without any loss of generality, let us assume this point to be $y(x^{k+1})$ corresponding to process No 1), sends the trial result to the master process. 
In turn, the master process selects a new trial point $x^{k+p+1}$ for the worker process according to the rules described below.
Note that in this case we will have a set of preimages of the trial points
\[
I_k = \left\{ x^{k+1},x^{k+2},...,x^{k+p} \right\},
\]
which the trials have been already started at but haven't been completed yet.

Step 1. Renumber set of preimages of the trial points 
\[
X_k = \left\{x^1, x^2,...,x^{k+p} \right\},
\]
containing all the preimages, which the trials either have been completed or are being performed at in the increasing order (by the lower index), so that
\[
0=x_1<x_2<...<x_{k+p}=1.
\]
Step 2. Compute the values 
\begin{gather*}
M_1=\max \left\{ \frac{ \left|z_i - z_{i-1} \right|}{(x_i-x_{i-1})^{1/N}} : x_{i-1} \notin I_k, x_i \notin I_k, 2\leq i\leq k+p \right\},\\
M_2=\max \left\{ \frac{ \left|z_{i+1} - z_{i-1} \right|}{(x_{i+1}-x_{i-1})^{1/N}} : x_i \in I_k, 2\leq i < k+p \right\},\\
M=\max\{M_1,M_2\},
\end{gather*}
where $z_i=\varphi(y(x_i))$ if $x_i \notin I_k, \; 1\leq i \leq k+p$. The values $z_i$ at the points $x_i \in I_k$ are undefined since the trials at the points $x_i \in I_k$ haven't been completed yet. If the value of $M$ equals 0, then set $M=1$.

Step 3. Juxtapose each interval $(x_{i-1},x_i), \; x_{i-1} \notin I_k, x_i \notin I_k, \; 2\leq i\leq k+p$ to a quantity $R(i)$, which is called a characteristic of the interval and is computed according to the formula
\begin{equation}\label{R}
R(i)=rM\Delta_i+\frac{(z_i-z_{i-1})^2}{rM\Delta_i}-2(z_i+z_{i-1}),
\end{equation}
where $\Delta_i=\left(x_i-x_{i-1}\right)^{1/N}$ and $r>1$ is the reliability parameter of the method.

Step 4. Select the interval $[x_{t-1},x_t]$, which the maximum characteristic corresponds to, i.e.
\[
R(t) = \max \left\{ R(i): \; x_{i-1} \notin I_k, x_i \notin I_k, \; 2\leq i\leq k+p \right\}.
\]

>>>>

\begin{document}
\mainmatter              % start of a contribution
% Решение обратных задач химической кинетки с помощью асинхронного алгоритма глобального поиска
% Решение обратных задач химической кинетки с помощью смешанного локально-глобального поискового алгоритма 
%\title{Solving the Inverse Problems of Chemical Kinetics Using the Asynchronous Global Optimization Algorithm}
\title{Kinetic Modeling of Isobutane Alkylation with Mixed C4 Olefins and Sulfuric Acid as a Catalyst Using the Asynchronous Global Optimization Algorithm}
\titlerunning{Kinetic Modeling of Isobutane Alkylation}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{
Irek Gubaydullin$^{1,2}$\and
Leniza Enikeeva$^{2,3}$\orcidID{0000-0003-4219-4870}
\and
Konstantin Barkalov$^4$ \letter \orcidID{0000-0001-5273-2471}
\and
Ilya Lebedev$^4$\orcidID{0000-0002-8736-0652} 
\and
Dmitry Silenko$^4$ 
}

@@

\institute{$^1$Institute of Petrochemistry and Catalysis -- Subdivision of Ufa Federal Research Centre of the RAS, Ufa, Russia\\$^2$Ufa State Petroleum Technological University, Ufa, Russia \\$^3$Novosibirsk State University, Novosibirsk, Russia\\$^4$Lobachevsky State University of Nizhny Novgorod, Nizhny Novgorod, Russia\\
\email{leniza.enikeeva@yandex.ru},
\email{konstantin.barkalov@itmm.unn.ru},
\email{ilya.lebedev@itmm.unn.ru}
}

@@

The paper considers the application of parallel computing technology to the simulation of a catalytic chemical reaction, which is widely used in the modern automobile industry to produce gasoline with a high octane number. As a chemical reaction, the process of alkylation of isobutane with mixed C4 olefins, catalyzed by sulfuric acid, is assumed. To simulate a chemical process, it is necessary to develop a kinetic model of the process, i.e., to determine the kinetic parameters. To do this, the inverse problem of chemical kinetics is solved; it predicts the values of the kinetic parameters based on laboratory data. From a mathematical point of view, the inverse problem of chemical kinetics is a global optimization problem. A parallel asynchronous information-statistical global search algorithm was used to solve it. The use of the asynchronous algorithm significantly reduced the search time to find the optimum. The found optimal parameters of the model made it possible to adequately simulate the process of alkylation of isobutane with mixed C4 olefins catalyzed by sulfuric acid.

%обновить ключевые слова
\keywords{Global optimization $\cdot$ Multi-extremal functions $\cdot$ Parallel computing $\cdot$ Chemical kinetics $\cdot$ Inverse problems }
\end{abstract}

@@

%Часть УГНТУ
Currently, there is a tendency to improve the environmental characteristics of automobile fuel while maintaining a high octane number. Sulfuric acid alkylation of isobutane with olefins makes it possible to obtain a high-octane component of gasoline with a minimum content of aromatic hydrocarbons. The alkylate, which is produced by alkylation of isobutane with C3 -- C5 olefins in the presence of strong acid, has the advantages of a high octane number, low vapor pressure, and zero content of olefins and aromatics, making it a desirable blending component for high-quality gasoline. Alkylates will continue to act as a desirable blending component for high-quality gasoline as the quality of gasoline continues to increase \cite{cao2019}. Therefore, it is a significant process for a modern refinery. To optimize the chemical process in industry, it is necessary to develop first its model, which in this case means building a mathematical model of the chemical process, and then its kinetic model, i.e., to numerically calculate the kinetic constants of the reaction.

%Стыковочный абзац
As a rule, it is impossible to find out the kinetic constants of reactions analytically. Therefore, there is a need in the development and application of numerical methods for finding the kinetic constants (see, e.g., \cite{Zaynullin2020,Enikeeva2020,Koledina2019,Enikeev2020,uskovfibrous,enikeeva2021,en13133393}). In this case, the quality criteria of the solution found (objective function) do not have an explicit analytical description, but enable an algorithmic representation and require considerable computational resources. Moreover, in the inverse problems of chemical kinetics, the objective function can be essentially multi-extremal, i.e., can have many local extrema along with the global one.

%Часть ННГУ
Numerical methods for solving such multi-extremal problems (global optimization methods) differ significantly from local search methods (see, e.g., \cite{Sergeyev2017,PaulaviciusZilinskas2014}). As a rule, local optimization methods cannot escape the local extremum attraction region and do not find the global optimum. At the same time, the use of model parameters corresponding to the found local solution may appear to be insufficient since the global solution can provide a considerable advantage over local ones.

The diversity of emerging global optimization problems entails various approaches to their solution. Methods for solving global optimization problems can be divided into two classes: metaheuristic and deterministic. Metaheuristic algorithms are usually based on the simulation of processes occurring in nature. Some examples of metaheuristic algorithms are simulated annealing, evolution and genetic algorithms, etc. (see, e.g., \cite{Battiti2009,Eiben2015}). Due to their relative simplicity, metaheuristic algorithms are more popular among researchers than deterministic methods. 
However, the problem solution found by the metaheuristic algorithm is, generally speaking, local and may be far from the global solution \cite{Kvasov2018}.

The possibility to construct deterministic global search methods different from grid search and metaheuristic methods is related to the availability and consideration of some {\it a priori} assumption on the properties of problem functions. Such assumptions play a key role in the development of efficient global optimization algorithms and serve as the main mathematical tool for estimating global solutions.

The assumption on limited relative variations of objective function values is one of the natural assumptions of the problem. Such an assumption is related to the ratio of the function increment to the respective increment of its argument, which is usually limited by some threshold defined by the limited energy of variations in the simulated system. In this case, the functions are known as Lipschitz ones, and the problem itself is called the Lipschitz global optimization problem.

This paper presents the results of applying parallel Lipschitz optimization methods for solving the inverse problems of chemical kinetics. The main part of the paper has the following structure. The description of the mathematical model of the investigated chemical reaction is presented in Section \ref{Sec_math_mod}. The formal statement of Lipschitz global optimization problems and the asynchronous parallel algorithm for solving them are described in Section \ref{Sec_GSA}. The results of the numerical solution of the inverse problem of chemical kinetics are discussed in Section \ref{Sec_Exp}.

@@

The initial conditions are $t = 0, c_1 = c_1^0; c_2=c_2^0; c_3 = 0; c_4 = 0; c_5= 0; c_6 = 0; c_7 = 0; c_8 = 0; c_9 = 0; c_{10} = 0; c_{11}=c_{11}^0; c_{12} = c_{12}^0$.
The corresponding species in equations (\ref{eq:one})--(\ref{eq:twelve}) are 1, iC4H8; 2, iC4, 3, iC4+; 4, TMPs+; 5, DMHs+; 6, TMPs; 7, DMHs; 8, HEs; 9, iCx+; 10, iCy=; 11, 2-C4H8; 12, 1-C4H8.

Information that represents a change in the concentrations of the reaction components over time at different temperatures is taken as experimental data from \cite{cao2019}; the data in Table \ref{table1} are presented as an example of experimental data at a temperature of 276.2 K.

\begin{table}
\caption{Experimental data}
\label{table1}
\begin{center}
\begin{tabular}{ccccccc}
\hline
time, min & 1 & 2 & 5 & 10 & 15 & 20 \\
\hline\rule{0pt}{12pt}
DMH & 0.12 & 0.11 & 0.1	& 0.1 &	0.095 &	0.09  \\
TMP & 0.54 & 0.65 & 0.69 & 0.69 & 0.7 & 0.705 \\[2pt]
\hline
\end{tabular}
\end{center}
\end{table}

@@

However, it is necessary to take into account the fact that the reaction rate constants $k_1, k_2, ..., k_{15}$ included in equations (\ref{eq:one})--(\ref{eq:twelve}) are parameters depending on the reaction temperature, this dependence is the Arrhenius equation and has the following form:

\begin{equation}
  k_i (T) = k_i^0 \exp \left(- \dfrac{E_i}{RT} \right),
  \label{eq:arren}
\end{equation}
where $k_i(T)$ is the constant of the $i$-th stage of the reaction rate, $k_i^0$ is the pre-exponential factor of the $i$-th reaction stage, $E_i$ is the activation energy, J/mol, $R$ is the universal gas constant, J $\cdot$ (K$\cdot$mol), $T$ is the temperature, K. Thus, to fully develop the kinetic model of the reaction, it is necessary to calculate the activation energies $E_i$ and the pre-exponential factors $k_i^0$ of all stages of the chemical reaction.  There are two formulations of the problems of searching for kinetic parameter data $E$ and $k^0$. The first one is to solve the inverse problem of selecting the kinetic parameters, included in equation (\ref{eq:arren}), which during the solution allow us to calculate all the reaction rate constants $k_i$. With the found rate constants of the reaction stages, the system of differential equations (\ref{eq:one})--(\ref{eq:twelve}) is solved, then the calculated concentrations are compared with the corresponding experimental data. Mathematically, this problem has the following formulation: it is necessary to minimize the following objective function

\begin{equation}
  F_1 = \sum_{i=1}^I \sum_{j=1}^J \sum_{k=1}^K \left| c_{ijk}^{exp} - c_{ijk}^{calc} \right| \longrightarrow \min,
  \label{eq:objective_func1}
\end{equation}
where $c_{ijk}^{exp}$ and $c_{ijk}^{calc}$ are the experimental and calculated values of the $k$-th observed component at the $i$-th experiment, respectively, $I$ is the number of observed temperatures for the reaction, ($I = 4$ in this case), $K$ is the number of experiments conducted at one temperature ($K = 3$), $M$ is the number of observed components of the reaction ($M = 2$).

The second formulation of the inverse problem of chemical kinetics implies the search for the rate constants of the stages $k_i$ included in the system (\ref{eq:one})--(\ref{eq:twelve}), separately for each temperature, then on the basis of Arrhenius equation~(\ref{eq:arren}), the activation energies $E_i$ and the pre-exponential multipliers $k_i^0$ are calculated using the least squares method. Mathematically, this statement of the problem coincides with equation (\ref{eq:objective_func1}), except for the first summation by temperatures:

@@

Thus, both formulations of the problem (\ref{eq:objective_func1}) and (\ref{eq:objective_func2}) are optimization problems, and the next section will describe the method used to solve these minimization problems.

@@

As mentioned above, from a formal point of view, we consider the inverse problem of chemical kinetics as a global optimization problem. 
In the specific problem under consideration, the values of the objective function are calculated by solving the stiff ODE system (\ref{eq:one})--(\ref{eq:twelve}). Since the right parts of the system are continuous functions with bounded derivatives, theoretically its solution will also be continuous and bounded. Therefore, discrepancy (\ref{eq:objective_func2}) will satisfy the Lipschitz condition with a priori unknown constant.

In the general form, the problem of the class specified above can be formulated mathematically as 
follows:
\begin{gather}
 \varphi^* = \varphi(y^\ast)=\min{\left\{\varphi(y):y\in D\right\}}, \label{problemN}\\
 D=\left\{y\in R^N: a_i\leq y_i \leq b_i, \;  1\leq i \leq N\right\} \label{D},
\end{gather}
where $a,b$ are the given vectors, $a,b\in R^N$, and the objective function $\varphi(y)$ satisfies the Lipschitz condition
\begin{equation}\label{Lip}
\left|\varphi(y_1)-\varphi(y_2)\right|\leq L\left\|y_1-y_2\right\|,\; y_1,y_2 \in D.
\end{equation}

The function $\varphi(y)$ is assumed to be multi-extremal and defined in the form of ``black box'' (i.e., in the form of some computing procedure, into the input of which the vector of parameters is supplied, and the corresponding function value is taken from the output). Moreover, each \textit{trial} (i.e., the computation of the function value at a point of the search domain) is assumed to be a time-consuming operation. 
As noted in Introduction, such a problem statement corresponds to the inverse problem of chemical kinetics completely.

Lipschitz condition (\ref{Lip}) can be utilized to estimate the global minimum of a function within an interval, and knowing the Lipschitz constant allows constructing global search algorithms and proving the convergence conditions for them (see, e.g., \cite{Strongin2000}).

The growth of computational costs with increasing the problem dimensionality is one of the main difficulties in solving multidimensional global optimization problems. Decreasing the number of trials at preserving the solution accuracy is possible by the complete utilization of some {\it a priori} assumptions on the objective function, which leads to adaptive serial optimization algorithms.

For example, the non-uniform space covering method \cite{Evtushenko2013} and the simplicial partitions method \cite{Zilinskas2010} are such methods. These approaches were successfully applied for the development of parallel optimization methods as well \cite{Evtushenko2009,Paulavicius2011}. 
Another adaptive approach to solving multidimensional problem (\ref{problemN}) is its reduction to a single one-dimensional problem or to several ones followed by the application of one-dimensional algorithms. 
Such a reduction can be made, for example, using the nested optimization scheme \cite{Grishagin2018} or Peano-Hilbert curves \cite{Barkalov2018}. 
The latter approach was used in the present work.

Using the continuous unambiguous mapping (Peano-Hilbert curve) $y(x)$ of the interval $[0,1]$ of the real axis on the hypercube $D$ from (\ref{D}), one can reduce multidimensional problem (\ref{problemN}) to a one-dimensional problem
\[
\varphi(y^\ast)=\varphi(y(x^\ast))=\min{\left\{\varphi(y(x)): x\in[0,1]\right\}},
\]
where the function $\varphi(y(x))$ will satisfy the uniform H{\"o}lder condition
\[
\left|\varphi(y(x_1))-\varphi(y(x_2))\right|\leq H\left|x_1-x_2\right|^{1/N}
\]
with the H{\"o}lder constant $H$ linked to the Lipschitz constant $L$ by the relation $ H=2 L \sqrt{N+3}$. 
The issues of the numerical construction of various approximations of the Peano-Hilbert curve were considered in \cite{Strongin2000,Sergeyev2013}.

So far, a search trial at some point $x'\in[0,1]$ will include first the construction of the image $y'=y(x')$ and then the computation of the value of the function $z'=\varphi(y')$.

@@

In the approach proposed, the parallelization scheme corresponds to the ``master/worker'' principle.
In the master process, the global search algorithm is executed, it accumulates search information, evaluates the Lipschitz constant for the objective function on its base, determines new trial points and distributes them among worker processes.

Worker processes receive the trial points from the master process, perform new trials at these points and send the trial results to the master process.

Let us assume that the master process computes one point of the next trial at each iteration and sends it to the worker process for executing the trial. 
At the same time, the execution of the trial by the worker process is a much more computationally expensive operation than the choice of a new trial point by the master that excludes idle worker processes. 
In this case (unlike synchronous parallel algorithms), the total number of trials executed by each worker process will depend on the computational costs of executing a particular trial and cannot be estimated in advance.

In the description of the parallel algorithm, let us assume that $p+1$ computational processes are at our disposal: one master process and $p$ worker ones.

At the beginning of the search, the master process (let us assume it to be Process No 0) initiates the parallel execution of $p$ trials at $p$ different points of the search domain. 
Two of these points are boundary, while the rest are internal, i.e., at the points $\{y(x^1), y(x^2), ...,y(x^p)\}$ where 
$x^1 = 0$, $x^p = 1$, $x^i\in(0,1), i=2,..., p-1$.

Now let us assume that $k\geq 0$ trials (in particular, $k$ can be equal to 0) are completed, and worker processes perform trials at the points $\{y(x^{k+1}), y(x^{k+2}), ...,y(x^{k+p})\}$.

Each worker process, having completed its trial at some point (without any loss of generality, let us assume this point to be $y(x^{k+1})$ corresponding to Process No 1), sends the trial result to the master process. 
In turn, the master process selects a new trial point $x^{k+p+1}$ for the worker process according to the rules described below.
Note that in this case we will have a set of preimages of the trial points
\[
I_k = \left\{ x^{k+1},x^{k+2},...,x^{k+p} \right\},
\]
at which the trials have already started, but have not yet been completed.

Step 1. Renumber the set of preimages of the trial points 
\[
X_k = \left\{x^1, x^2,...,x^{k+p} \right\},
\]
containing all preimages at which the trials either have been completed or underway in the increasing order (by the lower index) so that
\[
0=x_1<x_2<...<x_{k+p}=1.
\]
Step 2. Compute the values 
\begin{gather*}
M_1=\max \left\{ \frac{ \left|z_i - z_{i-1} \right|}{(x_i-x_{i-1})^{1/N}} : x_{i-1} \notin I_k, x_i \notin I_k, 2\leq i\leq k+p \right\},\\
M_2=\max \left\{ \frac{ \left|z_{i+1} - z_{i-1} \right|}{(x_{i+1}-x_{i-1})^{1/N}} : x_i \in I_k, 2\leq i < k+p \right\},\\
M=\max\{M_1,M_2\},
\end{gather*}
where $z_i=\varphi(y(x_i))$ if $x_i \notin I_k, \; 1\leq i \leq k+p$. The values $z_i$ at the points $x_i \in I_k$ are undefined since the trials at the points $x_i \in I_k$ have not yet been completed. If the value of $M$ equals 0, then set $M=1$.

Step 3. Juxtapose each interval $(x_{i-1},x_i), \; x_{i-1} \notin I_k, x_i \notin I_k, \; 2\leq i\leq k+p$ to the quantity $R(i)$, which is called the characteristic of the interval and is computed according to the formula
\begin{equation}\label{R}
R(i)=rM\Delta_i+\frac{(z_i-z_{i-1})^2}{rM\Delta_i}-2(z_i+z_{i-1}),
\end{equation}
where $\Delta_i=\left(x_i-x_{i-1}\right)^{1/N}$ and $r>1$ is the reliability parameter of the method.

Step 4. Select the interval $[x_{t-1},x_t]$, which the maximum characteristic corresponds to, i.e.,
\[
R(t) = \max \left\{ R(i): \; x_{i-1} \notin I_k, x_i \notin I_k, \; 2\leq i\leq k+p \right\}.
\]

Step 5. Define a new trial point $y^{k+p+1}=y(x^{k+p+1})$, the preimage of which is $x^{k+p+1} \in (x_{t-1},x_t)$ according to the formula
\[
x^{k+p+1} = \frac{x_{t}+x_{t-1}}{2} - \mathrm{sign}(z_{t}-z_{t-1})\frac{1}{2r}\left[\frac{\left|z_{t}-z_{t-1}\right|}{M}\right]^N.
\]

Upon computing the next trial point, the master process adds it to the set $I_k$ and sends it to the worker process, which initiates a new trial at this point.
